Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-13 Manuscript/Manuscript.Rmd
Add plotting functions for imputed data

# functions
miceadds::source.all(path = "../R")
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}

@@

We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by the sampling design. In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Add that we'll only discuss two levels?] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).

@@

```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
# missingness indicator plot (monotone vs non-monotone)
miss_ind <- tibble(
    rownr = 5:1,
    unit = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c("missing", rep("observed", 3), "missing")
  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))

# add text 
miss_ind$text <- ""
miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))

# plot
miss_ind %>% 
  mutate(name = factor(name, levels = unique(name))) %>% 
ggplot(aes(
  x = name,
  y = rownr,
  color = value,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text(show.legend = FALSE) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "") +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "")
# TODO: make it one plot with sporadic and systematic
# TODO: participant/response/*sample units*/records instead of obs on the y axis
# TODO: cluster 3
# TODO: add border/line
# TODO: remove unit
# TODO: add accolade

@@

<!-- - What types of multilevel methods are available? General overview of approaches, see @audi18 and @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods. -->

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well. Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. So we won't go into detail for the different methods (and equations). Refer to @meng94, an Audigier paper, and a paper by Grund on congeniality and random slopes. This paper is just a software tutorial. We'll keep it practical.]

@@

For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data.

We introduce three case studies to illustrate the workflow. In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model. In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data. In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.

@@

`popNCR` is a simulated dataset with pupils clustered in classes, where the number of cases $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

@@

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# plotting colors
plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))

# distributions
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
  geom_density() +
  scale_color_manual(values = plot_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}

# plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: legend: popteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal

# Feedback: 
# you only see the extreme ones in the positive sense, especially with 100 clusters
# alternatively add a quartile line around the density with geom_ribbon
```

@@

# further inspection of the imputations
densityplot(imp_ignored)

@@

# further inspection of the imputations
densityplot(imp_predictor)

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval}
# plot(imp_norm)
densityplot(imp_norm, ~popular)
ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
                    icc(popteach ~ as.factor(class), complete(imp_norm)), 
                    icc(texp ~ as.factor(class), complete(imp_norm)))
           )
ICCs
```

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval}
# plot(imp_pan)
densityplot(imp_pan, ~popular)
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
                    icc(popteach ~ as.factor(class), complete(imp_pan)), 
                    icc(texp ~ as.factor(class), complete(imp_pan)))
           )
ICCs
```

## Case study II: IMPACT

@@

## Pooling

- Analysis of scientific interest.

- Pooling using `mitml`.

- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorrelation)

>>>>

# functions
miceadds::source.all(path = "../R")
`%nin%` <- Negate(`%in%`)
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}

@@

We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by the sampling design. In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Alternatives to unit: participant/response/record.] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, no more?]

@@

```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
plot_md_set()
# TODO: add border/line
# TODO: remove unit
# TODO: add accolade

@@

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well. Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. So we won't go into detail for the different methods (and equations). Refer to @meng94, @audi18, and @grun18. This paper is just a software tutorial. We'll keep it practical.] We introduce three case studies to illustrate the workflow.

@@

For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data. In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model. In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data. In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.

@@

`popNCR` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

@@

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
# alternatively add a quartile line around the density with geom_ribbon
```

@@

# further inspection of the imputations
plot_box(imp_ignored, x = "popular", strip = TRUE)

@@

# further inspection of the imputations
plot_box(imp_predictor, x = "popular", strip = TRUE)

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm_2l <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval}
# plot(imp_norm)
plot_box(imp_norm_2l, x = "popular", strip = TRUE)

ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
           )
ICCs
```

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan_2l <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval}
# plot(imp_pan)
plot_box(imp_pan_2l, x = "popular", strip = TRUE)
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
           )
ICCs
```

# Case study II: IMPACT

@@

# Case study III: HIV

Toy example from [Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R).

### 2021-12-13 Manuscript/Manuscript.tex
Add plotting functions for imputed data

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field, clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'.\footnote{Add that we'll only discuss two
  levels?} For reasons of brevity, we only discuss clustering between
units, not within units (such as in timeseries or longitudinal data).

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations, although JOMO is available in
\pkg{mice} as well. Other useful packages for incomplete multilevel data
include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, an Audigier paper, and a paper by Grund on
  congeniality and random slopes. This paper is just a software
  tutorial. We'll keep it practical.}

@@

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data.

We introduce three case studies to illustrate the workflow. In the
\texttt{mice::popmis} data, we show the advantages of including the
multilevel structure of the data into the imputation model. In the
\texttt{metamisc::impact} data we'll show how to induce missingness and
solve it in real-world data. In the \texttt{GJRM::hiv} we provide novel
methodology\footnote{not really, the methods exist already, but how to
  show that this is something new and exciting?} for imputing MNAR
missingness according to the Heckman model.

@@

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
where the number of cases \(n = 2000\), and the number of clusters
\(N = 100\), on 7 variables:

@@

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.3338626 0.3726455
2 popteach  0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_norm)
R> densityplot(imp_norm, ~popular)
\end{CodeInput}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
+                     icc(popteach ~ as.factor(class), complete(imp_norm)), 
+                     icc(texp ~ as.factor(class), complete(imp_norm)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm
1  popular  0.3280070 0.3338626 0.3726455 0.3667167
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_pan)
R> densityplot(imp_pan, ~popular)
\end{CodeInput}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
+                     icc(popteach ~ as.factor(class), complete(imp_pan)), 
+                     icc(texp ~ as.factor(class), complete(imp_pan)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm       pan
1  popular  0.3280070 0.3338626 0.3726455 0.3667167 0.3719078
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{case-study-ii-impact}{%
\subsection{Case study II: IMPACT}\label{case-study-ii-impact}}

@@

\hypertarget{pooling}{%
\subsection{Pooling}\label{pooling}}

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorrelation)
\end{itemize}

>>>>

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field, clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'.\footnote{Alternatives to unit:
  participant/response/record.} For reasons of brevity, we only discuss
clustering between units, not within units (such as in timeseries or
longitudinal data).\footnote{Also, add that we'll only discuss two
  levels, no more?}

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations, although JOMO is available in
\pkg{mice} as well. Other useful packages for incomplete multilevel data
include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
  a software tutorial. We'll keep it practical.} We introduce three case
studies to illustrate the workflow.

@@

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
show the advantages of including the multilevel structure of the data
into the imputation model. In the \texttt{metamisc::impact} data we'll
show how to induce missingness and solve it in real-world data. In the
\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
  methods exist already, but how to show that this is something new and
  exciting?} for imputing MNAR missingness according to the Heckman
model.

@@

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
where the number of units \(n = 2000\), and the number of clusters
\(N = 100\), on 7 variables:

@@

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.3338626 0.3871020
2 popteach  0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm_2l <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_norm)
R> plot_box(imp_norm_2l, x = "popular", strip = TRUE)
\end{CodeInput}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_norm_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm
1  popular  0.3280070 0.3338626 0.3871020 0.3612830
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan_2l <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_pan)
R> plot_box(imp_pan_2l, x = "popular", strip = TRUE)
\end{CodeInput}

@@

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
+                     icc(texp ~ as.factor(class), complete(imp_pan_2l)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm       pan
1  popular  0.3280070 0.3338626 0.3871020 0.3612830 0.3745108
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{case-study-ii-impact}{%
\section{Case study II: IMPACT}\label{case-study-ii-impact}}

@@

\hypertarget{case-study-iii-hiv}{%
\section{Case study III: HIV}\label{case-study-iii-hiv}}

Toy example from
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
Github repo}.
